UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-06-27 ms/main.tex
tweak text

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller. Distributions of FEV1 by sex from selected Asian populations in figure~\ref{fig:supp_comparison_fev_chi_eur} are presented in figure~\ref{fig:supp_fev_ridgeplots}.

>>>>

Forced expiratory volume in 1 second (FEV1) (figures~\ref{fig:supp_ukbb_fev_f}, \ref{fig:supp_ukbb_fev_m}) also shows strong correlations with certain populations --- South Asian, African, and Caribbean --- having considerably lower measurements on average (see figures~\ref{fig:supp_box_fev_f} and \ref{fig:supp_box_fev_m} for boxplots and p-values). Notably, there appears to be a juncture in the admixture population, highlighted in figure~\ref{fig:supp_comparison_fev_afr}, where the distribution of FEV1 changes. This roughly corresponds to the transition from Black African/Caribbean individuals to those who identified having mixed backgrounds. Boxplots and statistical testing suggest that relative to White British populations, FEV1 values are significantly lower for Black African and Black Caribbean populations, but not for White and Black Caribbean and White and Black African populations. Figure~\ref{fig:supp_comparison_fev_chi_eur} suggests a difference in FEV1 between females who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. An unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 by 0.236 (p-value $4.902\times 10^{-11}$); adjusting for age, age$^2$, height, and sex shows a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though the sample of males is much smaller (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}).
